Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2023-11-06 docs/manuscript/paper.tex
version submitted to mnras

\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},

@@

Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}. At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$. This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation. The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$. The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}. For a fixed set of static parameters, the Kalman filter returns an estimate of the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error. We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline

@@

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
\end{equation}  
equivalent to a change of variables. We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e. Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.

@@

\section{SMBHB source parameters} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_new}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ is generally well-estimated (i.e. it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e. the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals). This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
	\label{fig:parameter_space}
\end{figure}
Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}. In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges. \newline

@@

\subsection{Identifiability}\label{sec:identif}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$). Nine coloured curves are plotted for nine random noise realizations. The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$. A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
	\label{fig:just_iota_h}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation. The red dot marks the maximum of $\log {\cal L}$. The colour scale is normalised with respect to the absolute value of the maximum. An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
	\label{fig:likelihood_surface}
\end{figure}
In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$. One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high. One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$. Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$. Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}. In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases. \newline

>>>>

\end{equation} where an overdot denotes a derivative with respect to $t$, and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},

@@

Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}. At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$. This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation. The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$. The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}. For a fixed set of static parameters, the Kalman filter returns an estimate of the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error. We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline

@@

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $dB(t)$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
\end{equation}  
equivalent to a change of variables. We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e. Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.

@@

\section{SMBHB source parameters} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ is generally well-estimated (i.e. it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e. the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals). This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
	\label{fig:parameter_space}
\end{figure}
Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}. In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges. \newline

@@

\subsection{Identifiability}\label{sec:identif}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$). Nine coloured curves are plotted for nine random noise realizations. The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$. A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
	\label{fig:just_iota_h}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation. The red dot marks the maximum of $\log {\cal L}$. The colour scale is normalised with respect to the absolute value of the maximum. An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
	\label{fig:likelihood_surface}
\end{figure}
In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$. One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section \ref{sec:pulsar_terms_drop}; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high. One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$. Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$. Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}. In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases. \newline
